Recurrent Deep Learning Models and its applications

### 2025-02-17 paper2/main.tex
paper update

\newpage
% ######################## Introduction ####################################
\section{Introduction}

With the rise of the Generative Artificial Intelligence, the development of AI has already made remarkable strides in processing sequential data. In understanding and producing sequential data. It has applications ranging from Natural Language Processing (NLP) to music composition to video generation. Especially NLP, has emerged as a pivotal field in artificial intelligence, enable machines to understand, interpret and generate in human readable format. Some famous Artificial Intellience assistence for example, Siri, Alexa and Bixby have shown the possibility. Everyone can communicate with those machines, which make the reasonable response back to user.\\[1ex]
Recurrent Neural Networks (RNNs) have been a foundational architecture in this domain, Unlike the traditional Artificial Neural Network, RNNs do not treat each input independently, RNNs handle each input by considering the information from previous inputs. Conceptually this architecture able to retain the information. Thus, this architecture is suitable for handling sequential data. Unfortunately, early RNNs had limitation in training of networks over long sequence. vanishing and exploding gradient problems significantly affect the training process of RNN \parencite{bengio1994learning}. Eliminating many practical applications of RNNs. After that, \parencite{hochreiter1997lstm} introduced Long Short-Term Memory (LSTM) networks and are responsible for the breakthrough in how to solve these challenges. Specificized gating mechanisms were introduced in LSTMs to regulate the flow of the information, minimize the vanishing gradient problem and learn the long-term dependencies. This advanced made RNNs much more performant on tasks like a language modeling, machine translation and speech recognition tasks.\\[1ex]
Further improvements were achieved with Gated Recurrent Units (GRUs) by \parencite{cho2014properties} which diminished the LSTM architecture's complexity, but still provided the same performance. GRUs performed comparably but used fewer parameters, making it computationally and more tractably trainable.\\[1ex]
Since the craze of AI has been revived by generative AI, natural language processing to time series prediction and speech recognition have once again aroused people's interest in RNN. This report aims to:
\begin{itemize}
    \item Explore the theoretical foundations of recurrent deep learning models.
    \item Investigate their diverse applications in solving sequential data tasks.
    \item Analyze their performance, strengths, and inherent limitations.
\end{itemize}
% ######################## Introduction ####################################
\newpage
% Background and Literature Review 
\section{Background and Literature Review}
% Evolution of Deep Learning and Recurrent Neural Networks 
\subsection{Evolution of Deep Learning and Recurrent Neural Networks}
In the past few decades, thank to the rapidly development of technology, the computing resource has a incredible increase. Thus, substantially deep learning architecture have improved, from simple architectures, which only able to capture simple information from data to sophisticated models that are able to learn complex, abstract representations. This was before the early neural networks like perceptrons and multilayer perceptrons (MLPs) laid the footwork of neural computation that first came in the picture, but were burdened by the lack of ability to model sequential dependencies. This however imposed a limit on the feed forward paradigm, which prompted the development of recurrent neural networks (RNNs) that extend the old stalactite of feed forward paradigm with cyclic connections. Through these connection, RNNs are capable to keep a hidden state that represents information over time steps thereby effectively capture temporal dynamics. RNNs have been a decisive step in the evolution of deep learning, as they are able to do tasks that require memory of previous events, including problems of natural language processing and time series modeling. Despite that, early RNNs models suffered from serious problems for example,  vanishing gradients and exploding gradients, which prevented these RNN models from learning long ranged dependencies. This stimulated the building of more refined architectures intended to side step these obstacles.

\newpage

@@

\newpage
\begin{longtable}{|c|c|c|}
    \hline
    \textbf{Notation} & \textbf{Meaning} & \textbf{Dimension}\\
    \hline
    $U$          & Weight matrix for input to hidden state       & $input\ size\times hidden\ unites$\\
    $W$          & Weight matrix for hidden to hidden state      & $hidden\ units\times hidden\ unites$\\
    $V$          & Weight matrix for hidden state to output state& $hidden\ units\times number\ of\ class$\\
    $x_t$        & Input vector at time t                        & $input\ size\times 1$\\
    $h_t$        & Hidden state output at time t                 & $hidden\ units\times 1$\\
    $b_h$        & Bias term for hidden state                    & $hidden\ units\times 1$\\
    $b_y$        & Bias term for output state                    & $number\ of\ class\times 1$\\
    $\hat{o}_y$  & Output at time t                              & $number\ of\ class\times 1$\\
    $\hat{y}_t$  & Output at time t                              & $hidden\ units\times 1$\\
    $\mathcal{L}$& Loss at time t                                & $scalar$\\
    \hline
    \caption{Unfolded RNN}
\end{longtable}
% Forward pass
\noindent \textbf{Forward Pass}
\\[1ex]
During the forward pass, the RNN processes the input sequence sequentially, computing hidden states and output at each timestep:
\begin{equation}
    h_t = f(U^Tx_{t}+W^Th_{t-1}+b_h)
\end{equation}
\begin{equation}
    \hat{y}_t = f(V^Th_t+b_y)
\end{equation}
\newline  % Computing the loss function
\noindent \textbf{Computing the loss function}
\\[1ex]
Assuming the loss is computed only at the final timestep t:
\begin{equation}
    \mathcal{L}_t = L(y_t, \hat{y}_t)
\end{equation}
In order to do backpropagation through time to tune the parameters in RNN, we need to calculate the partial derivative of loss function $\mathcal{L}$ with respect to the differently parameters.\\
\newline  % Backward pass using the chain rule
\noindent \textbf{Backward pass using the chain rule}
\\[1ex]
Using the chain rule for computing the gradient.\\
Partial derivative of loss function $\mathcal{L}$ with respect to $W$ (hidden to hidden state) at time 2. 
\begin{equation}
    \dfrac{\partial\mathcal{L}_2}{\partial W} = \dfrac{\partial\mathcal{L}_2}{\partial y_2} \cdot \dfrac{\partial y_2}{\partial h_2} \cdot \dfrac{\partial h_2}{\partial h_1} \cdot \dfrac{\partial h_1}{\partial W}
\end{equation}
By mathematic induction 
\begin{equation}
    \dfrac{\partial\mathcal{L}_t}{\partial W} = \dfrac{\partial\mathcal{L}_t}{\partial y_t} \cdot \dfrac{\partial y_t}{\partial h_t} \cdot (\sum_{i=1}^{t} \dfrac{\partial h_t}{\partial h_i} \cdot \dfrac{\partial h_i}{\partial W})
\end{equation}
Where 
\begin{equation}
    \dfrac{\partial h_t}{\partial h_i} = \prod_{j=i+1}^{t} \dfrac{\partial h_j}{\partial h_{j-1}}
\end{equation}

@@

\newpage
% Actication function
\subsubsection{Activation Function}
Activation functions, particularly the sigmoid function, are fundamental components of recurrent neural networks (RNNs). They transform input data into output data. A key property of these functions is their differentiability. Differentiability is crucial for the backpropagation through time (BPTT) algorithm, enabling the application of the chain rule during training. \\[1ex]
% Sigmoid Activation Function 
\textbf{Sigmoid activation function}
\\[1ex]
The main role of the sigmoid activation function is to normalize candidate values and convert the cell state to a hidden state when performing cell state updates. It limits the output between [0,1] because it has a smooth gradient, which is important for discovering long-range dependencies.\\[1ex]
\begin{equation}
    Sigmoid(x) = \frac{ 1 }{ 1+e^{-x} }
\end{equation}
\begin{equation}
    Sigmoid'(x) = Sigmoid(x)(1-Sigmoid(x))
\end{equation}
Below is the sigmoid function and its derivative.
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\textwidth]{../Pic/sigmoid.png} % Replace "example-image" with your image file
    % \caption{An example image.}
    \label{fig:sigmoid}
\end{figure}
\newline
$Domain(Sigmoid(x))=\mathbb{R},\hspace{1em} Codomain(Sigmoid(x))=(0,1)$\\
$Domain(Sigmoid'(x))=\mathbb{R},\hspace{1em} Codomain(Sigmoid'(x))=[0,0.5]$

% Hyperbolic tangent activation function
\newpage
\textbf{Hyperbolic tangent activation function}
\\[1ex]
The main role of the hyperbolic tangent (tanh) activation function is to normalize candidate values and convert the cell state to a hidden state when performing cell state updates. It limits the output between [-1,1] because it has a stable gradient, which is important for discovering long-range dependencies.
\newline
\begin{equation}
    tanh(x) = \frac{ e^{x} - e^{-x} }{e^{x} + e^{-x}}
\end{equation}
\begin{equation}
    tanh'(x) = 1 - tanh^{2}(x)
\end{equation}
Below is the Hyperbolic tangent activation function and its derivative.\\
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\textwidth]{../Pic/tanh.png} % Replace "example-image" with your image file
    % \caption{An example image.}
    \label{fig:tanh}
\end{figure}
\newline
$Domain(tanh(x))=\mathbb{R},\hspace{1em} Codomain(tanh(x))=[-1,1]$\\
$Domain(tanh'(x))=\mathbb{R},\hspace{1em} Codomain(tanh'(x))=[0,1]$

% Gradient vanishing and gradient exploring
\newpage
\subsubsection{Gradient vanishing and gradient exploring}
When training the RNN, BPTT was used to update the weight matrix. As the number of time steps increase, the problem of gradient instability of often encountered, and this problem is gradient vanishing and gradient explored \parencite{bengio1994learning}. 
\\[2ex]
% Vanishing Gradients
\textbf{Vanishing Gradients}
\\[1ex]
Generally, sigmoid activation function is used commonly in RNNs, has a maximum derivative of 0.25. When doing BPTT in long time steps, this multiplication results in exponentially diminishing gradients as the sequence length increases. Consequently, the shallow neural receive very small gradient updates, making it difficult to adjust the parameters effectively. This leads to the model struggling to learn long time dependencies. 
\\[1ex]
% Exploding Gradients
\textbf{Exploding Gradients}
\\[1ex]
When we are doing the feedforward and get super large value computed by loss function. Then when updating the parameters. The updates to the weights will also be large. Resulting in higher loss and larger gradients in the next iterations. This will lead to exploding gradients.
\\[1ex]
We have introduce the backpropagation through time. This is the method to update the parameters in RNNs. When calculating, for example, the partial derivative of loss function with respect to W. Assume the time $t$ go to infinity large. We will get this term. $ \prod_{j=i+1}^{t} \dfrac{\partial h_j}{\partial h_{j-1}} $, and it will lead to exponential problem. if $ \dfrac{\partial h_j}{\partial h_{j-1}} > 1 $. Then the product of all term will increase exponentially, then exploding gradients occur. On the contrary, if $\dfrac{\partial h_j}{\partial h_{j-1}} < 1$. Then the result will decrease exponentially, then vanishing gradients occur.
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\textwidth]{../Pic/exponential_growth.png} % Replace "example-image" with your image file
\end{figure}

\newpage

@@

\newpage
% Gated Recurrent Unit (GRU)
\subsubsection{Gated Recurrent Unit (GRU)}
The gated recurrent unit (GRU) was proposed by \parencite{cho2014properties} to make each recurrent unit to adaptively capture dependencies of different time scales. The GRU has 2 gates, update gate and reset gate. 
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.65\textwidth]{../Pic/gru.png} % Replace "example-image" with your image file
\end{figure}
Update gate:
\\[1ex]
The update gate determines how much of the past information should be retained in the current hidden state. 
\begin{equation}
    z_t = \sigma(W_z \cdot [ h_{t-1} , x_t ] + b_z )
\end{equation}
% Reset gate
Reset gate:
\\[1ex]
The reset gate is similar with the update gate, but the candidate hidden state is influenced by the reset gate. 
\begin{equation}
    r_t = \sigma( W_r \cdot [ h_{t-1} , x_t ] + b_r )
\end{equation}
Candidate hidden state:
\\[1ex]
The candidate hidden state combined with previous hidden state and current input to form the potential new information that can be added to the current hidden state.
\begin{equation}
    \widetilde{h_t} = tanh ( W_h \cdot [ r_t * h_{t-1} , x_t ] + b_h )
\end{equation}
Final hidden state
\\[1ex]
The final hidden state of the GRU at time t is a linear interpolation between the previous final hidden state and the candidate hidden state. 
\begin{equation}
    h_t = ( 1 - z_t ) * h_{t-1} + z_t * \widetilde{ h_t }
\end{equation}

@@

    % 3. Bias term
    \item Bias term
    \begin{itemize}
        \item Bias for update: $1 \times m$
        \item Bias for reset: $1 \times m$
        \item Bias for candidate hidden state: $1 \times m$
    \end{itemize}
\end{enumerate}
Total parameters: $3 \times ( n + m + 1 ) \times m$
\newpage

@@

\newpage
\subsubsection{Hidden Markov Model}
Before reviewing Hidden Markov Model (HMM), it is essential to understand what Markov models is, or Markov chains. Markov chains are fundamental models in probability theory and statistic, and it is a stochastic process describing a sequence of possible events in which the probability of each event depends only on the state attained in the previous event. This property known as the Markov property \parencite{roberts2004general, rabiner1986anintroduction}. In rigorous terms. Let the state space be defined as:
\begin{equation}
    S = \{ s_1, s_2, \dots, s_N \}.
\end{equation}
\begin{equation}
    P(s_{t+1} \mid s_t, s_{t-1}, s_{t-2}, \cdots, s_{1}) = P(s_{t+1} \mid s_t)
\end{equation}
The state transition probability is denoted by:
\begin{equation}
    P_{ij} = P(s_{t+1} = s_j \mid s_t = s_i).
\end{equation}

@@

\newpage

@@

\newpage

% Word representation 
\subsubsection{Word representation}
In RNN, the due to the architecture, the model can only handle vector, it is important that convert the word into machine readable format. By translating words into vectors, these representations capture semantic meanings, relationships and contexts. \parencite{mikolov2013efficient} proposed 2 model architectures for leaning distributed representations of words.
\\[1ex]
\textbf{Continuous Bag of Words model}
CBOW operates by taking context words around a target word, it is flexible to adjust the window size which mean that we can control how many words around our target word. Hence aggregating their embeddings and passing to hidden layer to produce a probability distribution. Capture the semantic meanings and the relationship more effectively. 
\begin{equation}
    Q = N \times D + D \times log_2(V)
\end{equation}
\\[1ex]
\textbf{Continuous Skip gram model}
\\[1ex]
The continuous skip gram model operates in opposite direction compare with CBOW. The model takes the surrounding word to predict the target word. In other word, treat the current word as an input to a log-linear classifier with continuous projection layer, and predict the probability distribution of surrounding words.
\begin{equation}
    Q = C \times (D + D \times log_2(V))
\end{equation}
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\textwidth]{../Pic/word_representation.png} % Replace "example-image" with your image file
\end{figure}

\newpage
% ######################## Background and Literature Review #########################

% ######################## Project Goals and Objectives ######################################
\subsection{Project Goals and Objectives}

% Project Goals
\subsubsection{Project Goals}
The primary goal of this report is to deeply review on RNN and how it can apply in natural language processing (NLP) tasks. And by comparing the performance of different models. this report aims to provide a detailed comparative analysis of the following RNN-based models:
\begin{itemize}
    \item Vanilla RNN
    \item Long Short-Term Memory (LSTM)
    \item Gated Recurrent Unit (GRU)
\end{itemize}

% Objectives
\subsubsection{Objectives}
\begin{enumerate}
    \item Implement Existing RNN Architectures\\
    Develop models using state-of-the-art deep learning framework, PyTorch, to implement each of the above RNNs model. This includes setting up the network layers, loss functions, and appropriate optimization methods.
    \item Apply the Models on a Benchmark NLP Dataset\\
    Utilize a selected NLP dataset to train and evaluate the performance of each RNN-based model. The dataset will be preprocessed (e.g., tokenization, padding, embedding initialization) to suit sequential data modeling.
    \item Compare Model Performances\\
    Compare different RNN-based models on metrics relevant to the target NLP taskquantitatively and qualitatively. Accuracy, F1, training time scores be used.
    \item Analyze the Impact of Architectural Differences
    Identify and discuss the strengths and limitations of each model architecture through both experimental results and theoretical insights. For example, compare how LSTM and GRU architectures mitigate vanishing gradients and exploding gradients using gating mechanisms relative to a vanilla RNN.
\end{enumerate}

\newpage
% ######################## Project Goals and Objectives ######################################

% ######################## Research Plan / Methodology ######################################
\subsection{Research Plan / Methodology}

>>>>

% ######################## Introduction ####################################
\section{Introduction}

With the rise of the Generative Artificial Intelligence, the development of AI has already made remarkable strides in processing sequential data. In understanding and producing sequential data. It has applications ranging from Natural Language Processing (NLP) to music composition to video generation. Especially NLP, has emerged as a pivotal field in artificial intelligence, enable machines to understand, interpret and generate in human readable format. Some famous Artificial Intellience assistence for example, Siri, Alexa and Bixby have shown the possibility. Everyone can communicate with those machines, which make the reasonable response back to user.\\[1ex]
Recurrent Neural Networks (RNNs) have been a foundational architecture in this domain, Unlike the traditional Artificial Neural Network, RNNs do not treat each input independently, RNNs handle each input by considering the information from previous inputs. Conceptually this architecture able to retain the information. Thus, this architecture is suitable for handling sequential data. Unfortunately, early RNNs had limitation in training of networks over long sequence. vanishing and exploding gradient problems significantly affect the training process of RNN \parencite{bengio1994learning}. Eliminating many practical applications of RNNs. After that, \parencite{hochreiter1997lstm} introduced Long Short-Term Memory (LSTM) networks and are responsible for the breakthrough in how to solve these challenges. Specificized gating mechanisms were introduced in LSTMs to regulate the flow of the information, minimize the vanishing gradient problem and learn the long-term dependencies. This advanced made RNNs much more performant on tasks like a language modeling, machine translation and speech recognition tasks.\\[1ex]
Further improvements were achieved with Gated Recurrent Units (GRUs) by \parencite{cho2014properties} which diminished the LSTM architecture's complexity, but still provided the same performance. GRUs performed comparably but used fewer parameters, making it computationally and more tractably trainable.\\[1ex]
Since the craze of AI has been revived by generative AI, natural language processing to time series prediction and speech recognition have once again aroused people's interest in RNN. This report aims to:
\begin{itemize}
    \item Explore the theoretical foundations of recurrent deep learning models.
    \item Investigate their diverse applications in solving sequential data tasks.
    \item Analyze their performance, strengths, and inherent limitations.
\end{itemize}
% ######################## Introduction ####################################

% ######################## Background and Literature Review #########################
% Background and Literature Review 
\section{Background and Literature Review}
% Evolution of Deep Learning and Recurrent Neural Networks 
\subsection{Evolution of Deep Learning and Recurrent Neural Networks}
In the past few decades, thank to the rapidly development of technology, the computing resource has a incredible increase. Thus, substantially deep learning architecture have improved, from simple architectures, which only able to capture simple information from data to sophisticated models that are able to learn complex, abstract representations. This was before the early neural networks like perceptrons and multilayer perceptrons (MLPs) laid the footwork of neural computation that first came in the picture, but were burdened by the lack of ability to model sequential dependencies. This however imposed a limit on the feed forward paradigm, which prompted the development of recurrent neural networks (RNNs) that extend the old stalactite of feed forward paradigm with cyclic connections. Through these connection, RNNs are capable to keep a hidden state that represents information over time steps thereby effectively capture temporal dynamics. RNNs have been a decisive step in the evolution of deep learning, as they are able to do tasks that require memory of previous events, including problems of natural language processing and time series modeling. Despite that, early RNNs models suffered from serious problems for example,  vanishing gradients and exploding gradients, which prevented these RNN models from learning long ranged dependencies. This stimulated the building of more refined architectures intended to side step these obstacles.

@@

\begin{longtable}{|c|c|c|}
    \hline
    \textbf{Notation} & \textbf{Meaning} & \textbf{Dimension}\\
    \hline
    $U$          & Weight matrix for input to hidden state       & $input\ size\times hidden\ unites$\\
    $W$          & Weight matrix for hidden to hidden state      & $hidden\ units\times hidden\ unites$\\
    $V$          & Weight matrix for hidden state to output state& $hidden\ units\times number\ of\ class$\\
    $x_t$        & Input vector at time t                        & $input\ size\times 1$\\
    $h_t$        & Hidden state output at time t                 & $hidden\ units\times 1$\\
    $b_h$        & Bias term for hidden state                    & $hidden\ units\times 1$\\
    $b_y$        & Bias term for output state                    & $number\ of\ class\times 1$\\
    $\hat{o}_y$  & Output at time t                              & $number\ of\ class\times 1$\\
    $\hat{y}_t$  & Output at time t                              & $hidden\ units\times 1$\\
    $\mathcal{L}$& Loss at time t                                & $scalar$\\
    \hline
    \caption{Unfolded RNN}
\end{longtable}
% Forward pass
\noindent \textbf{Forward Pass}
\\[1ex]
During the forward pass, the RNN processes the input sequence sequentially, computing hidden states and output at each timestep:
\begin{equation}
    h_t = f(U^Tx_{t}+W^Th_{t-1}+b_h)
\end{equation}
\begin{equation}
    \hat{y}_t = f(V^Th_t+b_y)
\end{equation}
\newline  % Computing the loss function
\noindent \textbf{Computing the loss function}
\\[1ex]
Assuming the loss is computed only at the final timestep t:
\begin{equation}
    \mathcal{L}_t = L(y_t, \hat{y}_t)
\end{equation}
In order to do backpropagation through time to tune the parameters in RNN, we need to calculate the partial derivative of loss function $\mathcal{L}$ with respect to the differently parameters.\\
\newline  % Backward pass using the chain rule
\noindent \textbf{Backward pass using the chain rule}
\\[1ex]
Using the chain rule for computing the gradient.\\
Partial derivative of loss function $\mathcal{L}$ with respect to $W$ (hidden to hidden state) at time 2. 
\begin{equation}
    \dfrac{\partial\mathcal{L}_2}{\partial W} = \dfrac{\partial\mathcal{L}_2}{\partial y_2} \cdot \dfrac{\partial y_2}{\partial h_2} \cdot \dfrac{\partial h_2}{\partial h_1} \cdot \dfrac{\partial h_1}{\partial W}
\end{equation}
By mathematic induction 
\begin{equation}
    \dfrac{\partial\mathcal{L}_t}{\partial W} = \dfrac{\partial\mathcal{L}_t}{\partial y_t} \cdot \dfrac{\partial y_t}{\partial h_t} \cdot (\sum_{i=1}^{t} \dfrac{\partial h_t}{\partial h_i} \cdot \dfrac{\partial h_i}{\partial W})
\end{equation}
Where 
\begin{equation}
    \dfrac{\partial h_t}{\partial h_i} = \prod_{j=i+1}^{t} \dfrac{\partial h_j}{\partial h_{j-1}}
\end{equation}

@@

% Actication function
\subsubsection{Activation Function}
Activation functions, particularly the sigmoid function, are fundamental components of recurrent neural networks (RNNs). They transform input data into output data. A key property of these functions is their differentiability. Differentiability is crucial for the backpropagation through time (BPTT) algorithm, enabling the application of the chain rule during training. \\[1ex]
% Sigmoid Activation Function 
\textbf{Sigmoid activation function}
\\[1ex]
The main role of the sigmoid activation function is to normalize candidate values and convert the cell state to a hidden state when performing cell state updates. It limits the output between [0,1] because it has a smooth gradient, which is important for discovering long-range dependencies.\\[1ex]
\begin{equation}
    Sigmoid(x) = \frac{ 1 }{ 1+e^{-x} }
\end{equation}
\begin{equation}
    Sigmoid'(x) = Sigmoid(x)(1-Sigmoid(x))
\end{equation}
Below is the sigmoid function and its derivative.
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\textwidth]{../Pic/sigmoid.png} % Replace "example-image" with your image file
    % \caption{An example image.}
    \label{fig:sigmoid}
\end{figure}
\newline
$Domain(Sigmoid(x))=\mathbb{R},\hspace{1em} Codomain(Sigmoid(x))=(0,1)$\\
$Domain(Sigmoid'(x))=\mathbb{R},\hspace{1em} Codomain(Sigmoid'(x))=[0,0.5]$

% Hyperbolic tangent activation function
\textbf{Hyperbolic tangent activation function}
\\[1ex]
The main role of the hyperbolic tangent (tanh) activation function is to normalize candidate values and convert the cell state to a hidden state when performing cell state updates. It limits the output between [-1,1] because it has a stable gradient, which is important for discovering long-range dependencies.
\newline
\begin{equation}
    tanh(x) = \frac{ e^{x} - e^{-x} }{e^{x} + e^{-x}}
\end{equation}
\begin{equation}
    tanh'(x) = 1 - tanh^{2}(x)
\end{equation}
Below is the Hyperbolic tangent activation function and its derivative.\\
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\textwidth]{../Pic/tanh.png} % Replace "example-image" with your image file
    % \caption{An example image.}
    \label{fig:tanh}
\end{figure}
\newline
$Domain(tanh(x))=\mathbb{R},\hspace{1em} Codomain(tanh(x))=[-1,1]$\\
$Domain(tanh'(x))=\mathbb{R},\hspace{1em} Codomain(tanh'(x))=[0,1]$

% Gradient vanishing and gradient exploring
\subsubsection{Gradient vanishing and gradient exploring}
When training the RNN, BPTT was used to update the weight matrix. As the number of time steps increase, the problem of gradient instability of often encountered, and this problem is gradient vanishing and gradient exploring \parencite{bengio1994learning}. 
\\[2ex]
% Vanishing Gradients
\textbf{Vanishing Gradients}
\\[1ex]
Generally, sigmoid activation function is used commonly in RNNs, has a maximum derivative of 0.25. When doing BPTT in long time steps, this multiplication results in exponentially diminishing gradients as the sequence length increases. Consequently, the shallow neural receive very small gradient updates, making it difficult to adjust the parameters effectively. This leads to the model struggling to learn long time dependencies. 
\\[1ex]
% Exploding Gradients
\textbf{Exploding Gradients}
\\[1ex]
When we are doing the feedforward and get super large value computed by loss function. Then when updating the parameters. The updates to the weights will also be large. Resulting in higher loss and larger gradients in the next iterations. This will lead to exploding gradients.
\\[1ex]
We have introduce the backpropagation through time. This is the method to update the parameters in RNNs. When calculating, for example, the partial derivative of loss function with respect to W. Assume the time $t$ go to infinity large. We will get this term. $ \prod_{j=i+1}^{t} \dfrac{\partial h_j}{\partial h_{j-1}} $, and it will lead to exponential problem. if $ \dfrac{\partial h_j}{\partial h_{j-1}} > 1 $. Then the product of all term will increase exponentially, then exploding gradients occur. On the contrary, if $\dfrac{\partial h_j}{\partial h_{j-1}} < 1$. Then the result will decrease exponentially, then vanishing gradients occur.
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\textwidth]{../Pic/exponential_growth.png} % Replace "example-image" with your image file
\end{figure}

@@

% Gated Recurrent Unit (GRU)
\subsubsection{Gated Recurrent Unit (GRU)}
The gated recurrent unit (GRU) was proposed by \parencite{cho2014properties} to make each recurrent unit to adaptively capture dependencies of different time scales. The GRU has 2 gates, update gate and reset gate. 
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.65\textwidth]{../Pic/gru.png} % Replace "example-image" with your image file
\end{figure}
Update gate:
\\[1ex]
The update gate determines how much of the past information should be retained in the current hidden state. 
\begin{equation}
    z_t = \sigma(W_z \cdot [ h_{t-1} , x_t ] + b_z )
\end{equation}
% Reset gate
Reset gate:
\\[1ex]
The reset gate is similar with the update gate, but the candidate hidden state is influenced by the reset gate. 
\begin{equation}
    r_t = \sigma( W_r \cdot [ h_{t-1} , x_t ] + b_r )
\end{equation}
Candidate hidden state:
\\[1ex]
The candidate hidden state combined with previous hidden state and current input to form the potential new information that can be added to the current hidden state.
\begin{equation}
    \widetilde{h_t} = tanh ( W_h \cdot [ r_t * h_{t-1} , x_t ] + b_h )
\end{equation}
Final hidden state
\\[1ex]
The final hidden state of the GRU at time t is a linear interpolation between the previous final hidden state and the candidate hidden state. 
\begin{equation}
    h_t = ( 1 - z_t ) * h_{t-1} + z_t * \widetilde{ h_t }
\end{equation}

@@

    % 3. Bias term
    \item Bias term
    \begin{itemize}
        \item Bias for update: $1 \times m$
        \item Bias for reset: $1 \times m$
        \item Bias for candidate hidden state: $1 \times m$
    \end{itemize}
\end{enumerate}
Total parameters: $3 \times ( n + m + 1 ) \times m$

@@

\subsubsection{Hidden Markov Model}
Before reviewing Hidden Markov Model (HMM), it is essential to understand what Markov models is, or Markov chains. Markov chains are fundamental models in probability theory and statistic, and it is a stochastic process describing a sequence of possible events in which the probability of each event depends only on the state attained in the previous event. This property known as the Markov property \parencite{roberts2004general, rabiner1986anintroduction}. In rigorous terms. Let the state space be defined as:
\begin{equation}
    S = \{ s_1, s_2, \dots, s_N \}.
\end{equation}
\begin{equation}
    P(s_{t+1} \mid s_t, s_{t-1}, s_{t-2}, \cdots, s_{1}) = P(s_{t+1} \mid s_t)
\end{equation}
The state transition probability is denoted by:
\begin{equation}
    P_{ij} = P(s_{t+1} = s_j \mid s_t = s_i).
\end{equation}

@@

@@

% Word representation 
\subsubsection{Word representation}
In RNN, the due to the architecture, the model can only handle vector, it is important that convert the word into machine readable format. By translating words into vectors, these representations capture semantic meanings, relationships and contexts. \parencite{mikolov2013efficient} proposed 2 model architectures for leaning distributed representations of words.
\\[1ex]
\textbf{Continuous Bag of Words model}
\\[1ex]
CBOW operates by taking context words around a target word, it is flexible to adjust the window size which mean that we can control how many words around our target word. Hence aggregating their embeddings and passing to hidden layer to produce a probability distribution. Capture the semantic meanings and the relationship more effectively. 
\begin{equation}
    Q = N \times D + D \times log_2(V)
\end{equation}
\\[1ex]
\textbf{Continuous Skip gram model}
\\[1ex]
The continuous skip gram model operates in opposite direction compare with CBOW. The model takes the surrounding word to predict the target word. In other word, treat the current word as an input to a log-linear classifier with continuous projection layer, and predict the probability distribution of surrounding words.
\begin{equation}
    Q = C \times (D + D \times log_2(V))
\end{equation}
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\textwidth]{../Pic/word_representation.png} % Replace "example-image" with your image file
\end{figure}

% ######################## Project Goals and Objectives ######################################
\section{Project Goals and Objectives}

% Project Goals
\subsection{Project Goals}
The primary goal of this report is to deeply review on RNN and how it can apply in natural language processing (NLP) tasks. And by comparing the performance of different models. this report aims to provide a detailed comparative analysis of the following RNN-based models:
\begin{itemize}
    \item Vanilla RNN
    \item Long Short-Term Memory (LSTM)
    \item Gated Recurrent Unit (GRU)
\end{itemize}

% Objectives
\subsection{Objectives}
\subsubsection{Implement Existing RNN Architectures}
Develop models using state-of-the-art deep learning framework, PyTorch, to implement each of the above RNNs model. This includes setting up the network layers, loss functions, and appropriate optimization methods.
\subsubsection{Apply the Models on a Benchmark NLP Dataset}
Utilize a selected NLP dataset to train and evaluate the performance of each RNN-based model. The dataset will be preprocessed (e.g., tokenization, padding, embedding initialization) to suit sequential data modeling.
\subsubsection{Compare Model Performances}
Compare different RNN-based models on metrics relevant to the target NLP taskquantitatively and qualitatively. Accuracy, F1, training time scores be used.
\subsubsection{Analyze the Impact of Architectural Differences}
Identify and discuss the strengths and limitations of each model architecture through both experimental results and theoretical insights. For example, compare how LSTM and GRU architectures mitigate vanishing gradients and exploding gradients using gating mechanisms relative to a vanilla RNN.

% ######################## Project Goals and Objectives ######################################

% ######################## Research Plan / Methodology ######################################
\section{Research Plan / Methodology}
To achieve the project goals and objectives, this report will follow the step as below:
\subsection{Literature Review}
Since this report require some knowledge related on RNN and NLP. So it is necessary to review the foundemental of those area. Different architecture of RNNs-based model for instance, vanilla RNN, LSTM, GRU. And understand the limitation of RNN, vanishing/exploding gradients. Also, it is essential to understand natural language processing (NLP).  
\subsection{Data Collection and Preprocessing}
In this report, ACL-IMDB \parencite{maas-EtAl:2011:ACL-HLT2011} was experiment. This is a dataset for binary sentiment classification containing substantially more data than previous benchmark datasets. This dataset provide a set of 25,000 highly polar movie reviews for training, and 25,000 for testing. I preporcess the dataset by text cleaning and tokenization.
\subsection{Model Architecture}
